---
title: "Assignment 7 - Repeated measure"
author: "Md Shakir Moazzem"  
format:
  html:
    code-fold: true 
    embed-resources: true
    toc: true
    number-sections: true
    theme: cerulean
---

# Introduction  
This assignment will ask you to conduct a repeated measure analysis. 

For that, you will first need to watch (and code along with) the following repeated measure video:

- repeated measure lecture: https://youtu.be/-ANqcQhB--A?si=7YeeYqDgcLsj5erW&t=4268 (starts at 01:11:08, watch till the end, about 50 min)

- repeated measure lab part 1: https://youtu.be/GY2e-9M5bFM?si=TIw5F4EF2m9Dx5uK (watch full video, ~1h 50 min)  

- repeated measure lab part 2: https://youtu.be/F3vg89ymLeM?si=Zv-RPIdS8mig4EMb
(watch full video, ~2h)

- The partial script for the video above is in our GitHub repository, 04-classcode, **03-11_repeatedmeasure_partial.qmd**.  

- The data set for this exercise is in oir GitHub repository, 03-classcode, **wheat_nk_balkh_rm.csv**.


**AFTER watching the video and developing the code in class**, then start working on this HW assignment.  

# Instructions  

  - Create on your GitHub a repo named `06_rm`.  
  - Clone it to your computer, create the necessary subfolders.
  - Move the script **03-11_repeatedmeasure_partial.qmd** into the `code` subfolder.
  - Move this script into the `code` subfolder  
  - Move the **wheat_nk_balkh_rm.csv** data set into the `data` subfolder  
  - On this script, fill your name under the `title` section of the YAML  
  
# Goal of this assignment  
The goal is this assignment is for you to **learn about and conduct a full repeated measure analysis**.

Once you finish coding along with the videos, answer the following questions.

# Questions  
In class, we decided to use AIC as the metric to select the best model, and this determined the best model to be the one with a general error correlation matrix structure. If we had used BIC as the metric, then the model selected would have been the default model. Here, you are asked to use BIC and thus select the default model as the best, and perform all the steps after model selection. You may copy and paste the class script, and adapt the copied version to implement these changes.  

# Setup and wrangling  
```{r}
#| message: false
#| warning: false
library(tidyverse) # for data wrangling and plotting
library(car) # for Anova function
library(lme4)
library(broom) # for model residuals extraction
library(emmeans) # for model mean extraction
library(multcomp) # for pairwise comparison letter display
library(nlme) # for repeated measure correlation structures 

#We use "lme4" package for mixed effects ANOVA models.
#We use "nlme" package for repeated measures ANOVA models.

```

Reading data and doing some light wrangling.  
```{r}
#| message: false

#In this data frame, "dap" stands for "days after planting" with 4 levels: 120, 130, 140, 150

rm_rcbd_df <- read_csv("../data/wheat_nk_balkh_rm.csv") %>%
  mutate(across(trt:splot, ~factor(.))) %>%
  mutate(fdap = factor(dap)) #fdap = dap as factor #because we want to have "dap" as both numeric and categorical columns

#1st argument in across() function: select all columns between "trt" and "splot" #2nd argument: ~factor(.) here . is a place holder into which all selected columns (trt:splot) are passed

rm_rcbd_df

```

Since we have been instructed to select the default model as the best model based on the BIC, I am going to skip all the intermediary steps and going straight to perform all the steps **after** model selection.

# Model - Default  Model

```{r}
mod1_default <- lme(yield_kgha ~ krate_kgha*nrate_kgha*fdap, #Specifying fixed term/effect
                    random = ~ 1 | block/krate_kgha/nrate_kgha, #Specifying random term/effect
                    data = rm_rcbd_df #Specifying where the data is coming from
                    )

rm_rcbd_df

mod1_default
```

# Model - General Model

Compound symmetry is the simplest:  correlation = corCompSymm(form = ~ dap)
General is the most complex: 

```{r}

mod3_gen <- lme(
  yield_kgha ~ krate_kgha*nrate_kgha*fdap,
  random = ~ 1|block/krate_kgha/nrate_kgha,
  correlation = corSymm(),
  data = rm_rcbd_df)

mod3_gen

```


## Q1: Include here the residual assessment of the default model. Include each plot and its interpretation. How do these residuals from the default model visually compare to those from the general model?  

# Checking residuals for the general model  

```{r residual general model}

library(broom.mixed)

mod3_gen_resid <- augment(mod3_gen) %>%
  mutate(.stdresid=resid(mod3_gen, 
                         type="pearson", 
                         scaled=T)) # resid() = Standardized residual (not studentized residual)

mod3_gen_resid
```

# Checking residuals for the default model 

```{r residual default model}

library(broom.mixed)

mod1_default_resid <- augment(mod1_default) %>%
  mutate(.stdresid=resid(mod1_default, 
                         type="pearson", 
                         scaled=T)) # resid() = Standardized residual (not studentized residual)

mod1_default_resid
```

### Block random effects are iid ~ N(0,var_a)  

```{r general model: Block random effects}
ranef(mod3_gen)[[1]] %>% 
  ggplot(aes(sample=`(Intercept)`))+
  stat_qq(  shape = 21,
            fill = "purple", 
            size = 3,
            alpha = .7
  )+
  stat_qq_line()+
  labs(x = "Theoretical quantile",
       y = "Sample quantile",
       title = "General model: Blocks Random Effects QQ plot")+
  theme_bw()

```


```{r default model: Block random effects}
ranef(mod1_default)[[1]] %>% 
  ggplot(aes(sample=`(Intercept)`))+
  stat_qq(  shape = 21,
            fill = "purple", 
            size = 3,
            alpha = .7
  )+
  stat_qq_line()+
  labs(x = "Theoretical quantile",
       y = "Sample quantile",
       title = "Default model: Blocks Random Effects QQ plot")+
  theme_bw()

```


### Block:krate random effects are iid ~ N(0,var_b) [random effect associated with potassium rate nested within block]

```{r general model: Block:krate random effects}

ranef(mod3_gen)[[2]] %>% 
  ggplot(aes(sample=`(Intercept)`))+
  stat_qq(  shape = 21,
            fill = "purple", 
            size = 3,
            alpha = .7
  )+
  stat_qq_line()+
  labs(x = "Theoretical quantile",
       y = "Sample quantile",
       title = "General model: Block*krate Random Effects QQ plot")+
  theme_bw()


```


```{r default model: Block:krate random effects}
ranef(mod1_default)[[2]] %>% 
  ggplot(aes(sample=`(Intercept)`))+
  stat_qq(  shape = 21,
            fill = "purple", 
            size = 3,
            alpha = .7
  )+
  stat_qq_line()+
  labs(x = "Theoretical quantile",
       y = "Sample quantile",
       title = "Default model: Block*krate Random Effects QQ plot")+
  theme_bw()

```

### Block:krate:nrate random effects are iid ~ N(0,var_c)  [random effect: nitrogen rate inside potassium rate inside block]


```{r general model: Block:krate:nrate Random Effects QQ plot}

ranef(mod3_gen)[[3]] %>% 
  ggplot(aes(sample=`(Intercept)`))+
  stat_qq(  shape = 21,
            fill = "purple", 
            size = 3,
            alpha = .7
  )+
  stat_qq_line()+
  labs(x = "Theoretical quantile",
       y = "Sample quantile",
       title = "General model: Block*krate*nrate Random Effects QQ plot")+
  theme_bw()

```


```{r default model: Block:krate:nrate random effects QQ plot}

ranef(mod1_default)[[3]] %>% 
  ggplot(aes(sample=`(Intercept)`))+
  stat_qq(  shape = 21,
            fill = "purple", 
            size = 3,
            alpha = .7
  )+
  stat_qq_line()+
  labs(x = "Theoretical quantile",
       y = "Sample quantile",
       title = "Default model: Block*krate*nrate Random Effects QQ plot")+
  theme_bw()

```
### Within-group errors are iid ~ N(0, var_e)  

```{r general model: fitted vs. residual plot for Within-group errors, warning=FALSE, message=FALSE}

ggplot(mod3_gen_resid, aes(x=.fitted, y=.stdresid))+
  geom_hline(yintercept = 0, color="red")+
  geom_point(shape = 21,
             fill = "purple", 
             size = 3,
             alpha = .7)+
  geom_smooth()+
  geom_hline(yintercept = c(-3,3), color = "red")+
  labs(y = "Standardized Residuals",
       x = "Fitted Values",
       title = "General model: Fitted vs. Residual Plot for Within-group Errors")+
  theme_bw()

```

```{r default model: fitted vs. residual plot for Within-group errors, warning=FALSE, message=FALSE}

ggplot(mod1_default_resid, aes(x=.fitted, y=.stdresid))+
  geom_hline(yintercept = 0, color="red")+
  geom_point(shape = 21,
             fill = "purple", 
             size = 3,
             alpha = .7)+
  geom_smooth()+
  geom_hline(yintercept = c(-3,3), color = "red")+
  labs(y = "Standardized Residuals",
       x = "Fitted Values",
       title = "Default model: Fitted vs. Residual Plot for Within-group Errors")+
  theme_bw()

```


```{r General Model: QQ Plot for Within-group Errors, warning=FALSE, message=FALSE}

ggplot(mod3_gen_resid, aes(sample=.stdresid))+
  stat_qq(  shape = 21,
            fill = "purple", 
            size = 3,
            alpha = .7
  )+
  stat_qq_line()+
  labs(x = "Theoretical quantile",
       y = "Sample quantile",
       title = "General Model: QQ Plot for Within-group Errors")+
  theme_bw()

```

```{r Default Model: QQ Plot for Within-group Errors, warning=FALSE, message=FALSE}

ggplot(mod1_default_resid, aes(sample=.stdresid))+
  stat_qq(  shape = 21,
            fill = "purple", 
            size = 3,
            alpha = .7
  )+
  stat_qq_line()+
  labs(x = "Theoretical quantile",
       y = "Sample quantile",
       title = "Default Model: QQ Plot for Within-group Errors")+
  theme_bw()

```



## Q2: Include here the ANOVA table from the default model. How does it compare with the ANOVA table from the general model?  

### ANOVA Table for General Model  

```{r anova general model, warning=FALSE, message=FALSE}

Anova(mod3_gen, type = 3)

```


### ANOVA Table for Default Model  

```{r anova default model, warning=FALSE, message=FALSE}

Anova(mod1_default, type = 3)

```

For the "mod1_default" i.e., default model, the 3-way "krate_kgha:nrate_kgha:fdap" interaction (*p*-value = 0.06951) is not significant at Alpha = 0.05 significance level, rather the 2-way "krate_kgha:nrate_kgha" interaction (*p*-value = 1.047e-06) and the "fdap" main effect (*p*-value < 2.2e-16) are significant at Alpha = 0.001 significance level. Hence, we would mention 2 different pairwise letter separation plots/tables/interpretation paragraphs: one for the 2-way "krate_kgha:nrate_kgha" interaction, and another plot/table/interpretation paragraph for the "fdap" main effect (because both "krate_kgha:nrate_kgha" and "fdap"  were significant at Alpha = 0.001 significance level).


In contrast, the "mod3_gen" i.e., general model had the 3-way "krate_kgha:nrate_kgha:fdap" interaction (*p*-value = 7.576e-05) to be highly significant at Alpha = 0.001 significance level. Hence, we would extract means and mention pairwise letter separation plots/tables/interpretation paragraphs either to compare different days after planting (fdap) within all combinations of nitrogen rate and potassium rates, or to compare all combinations of nitrogen and potassium rates within each day after planting (fdap).

# Submitting your work  
Once you have developed all the code and answers, make sure to Render this quarto file.  

**Notes on rendering**:  

- Make sure to render your work and inspect how the final html look like.  
- If it does not look professional for whatever reason, then fix the issue, re-render it, recheck.  
- Only send me your work once your html file looks professional.  
  - **DO NOT** delete the file's heading levels (# and ##). They set up the proper heading 1 and 2 levels, and I use them to guide my grading.  
  - If a given chunk is also outputting warnings or messages, inhibit this behavior by changing the chunk options `message` and `warning` to `FALSE` (I'll demo this in a moment).  
  
  - If, after rendered, 2 lines of text are connected and you wish to "break line" between them, add 2 extra spaces after the first one (I'lld demo this in a moment).  

After rendering, an .html file will be created on your `code` folder.  

Rename this file to `Assignment-07-rm-LASTNAME.html`.    

For ex., mine would be `Assignment-07-rm-Bastos.html`.

Submit your work by uploading the **html** file to **eLC Assigment #7 - Repeated measure** by Mar 25 11:59 pm.  
  




